The Lipid Interactome

### 2025-02-11 Background/MultifunctionalLipidProbesOverview.qmd
Incorporating minor changes after weekly meeting

At its core, biology is a complex network of chemical interactions. Every biological process—from metabolism to signaling—depends on the physical interactions between molecules, including proteins, nucleic acids, and lipids. While proteins have long been the primary focus of interaction studies, lipids play an equally crucial yet often underappreciated role in cellular function. Therefore, there is a need for detailed knowledge of lipid-protein interactions (the lipid interactome), in healthy and diseased cells and tissue.

@@

To overcome these challenges, researchers have developed synthetic analogs of native lipids. These modified lipids allow for precise investigation of lipid function while preserving their essential biological properties. The key modifications used in this approach include a photo-crosslinking group (usually a diazirine), an alkyne group for click chemistry, and a photo-removable (cage) group for preventing premature probe metabolism.:

@@

##### A Versatile Bioorthogonal Handle

@@

The power of coumarin-caged lipids has been demonstrated in studies examining intracellular lipid signaling. For example, in 2015, Högliner et al. used a caged sphingosine analog to investigate the role of sphingosine in calcium signaling. They found that upon photoactivation of the caged sphingosine, lysosomal calcium release was triggered, revealing a crucial link between sphingosine metabolism and intracellular calcium dynamics [@hoglingerIntracellularSphingosineReleases2015a]. This discovery significantly expanded the understanding of the sphingosine signaling cascade and its downstream effects.

@@

<!--
## Photoactivatable & Clickable Probes / Bifunctional Lipid Probes

 -   The conceptual precursor to the so-called trifunctional probes, these typically include only the diazirine and the terminal alkyne moieties.

References to include:

-   [@haberkantFatFabulousBifunctional2014] - Haberkant et al. 2014

## Trifunctional Lipid Probes
-->

>>>>

At its core, biology is a complex network of chemical interactions. Every biological process—from metabolism to signaling—depends on the physical interactions between molecules, including proteins, nucleic acids, and lipids. While proteins have long been the primary focus of interaction studies, lipids play an equally crucial yet often under-appreciated role in cellular function. Therefore, there is a need for detailed knowledge of lipid-protein interactions (the lipid interactome), in healthy and diseased cells and tissue.

@@

To overcome these challenges, researchers have developed synthetic analogs of native lipids. These modified lipids allow for precise investigation of lipid function while preserving their essential biological properties. The key modifications used in this approach include a photo-crosslinking group (usually a diazirine), an alkyne group for click chemistry, and a photo-removable (cage) group for preventing premature probe metabolism.

@@

##### A versatile bioorthogonal handle

@@

The power of coumarin-caged lipids has been demonstrated in studies examining intracellular lipid signaling. For example, in 2015, Höglinger et al. used a caged sphingosine analog to investigate the role of sphingosine in calcium signaling. They found that upon photoactivation of the caged sphingosine, lysosomal calcium release was triggered, revealing a crucial link between sphingosine metabolism and intracellular calcium dynamics [@hoglingerIntracellularSphingosineReleases2015a]. This discovery significantly expanded the understanding of the sphingosine signaling cascade and its downstream effects.

@@

### 2025-02-11 BioinformaticsApplicationNote/GG_2025_applicationNote.qmd
Incorporating minor changes after weekly meeting

copyright:
  holder: "Gaelen Guzman"
  year: 2025
authors:
  - name: "Gaelen Guzman"
    affiliations:
      - name:  "Oregon Health & Science University"
        address: 3222 SW Research Dr
        city: Portland
        state: OR
        postal-code: 97239
    orcid: 0000-0002-7696-6034
    corresponding: true
    email: gaelen.guzman@lipidinteractome.org
  - name: "Carsten Schultz"
    affiliations:
      - name:  "Oregon Health & Science University"
        address: 3222 SW Research Dr
        city: Portland
        state: OR
        postal-code: 97239
    orcid: 0000-0002-5824-2171
  - name: "Fikadu Tafesse"
    affiliations:
      - name:  "Oregon Health & Science University"
        address: 3222 SW Research Dr
        city: Portland
        state: OR
        postal-code: 97239
    orcid: 0000-0002-8575-4164
format:
  html:
    embed-resources: true
  docx:
    fig-height: 5
    reference-doc: /Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/Resources/custom-reference-doc.docx
page-navigation: false
---

@@

### Data Submission

If you would like to submit a relevant dataset, see our [Data Submission Page](https://www.lipidinteractome.org/DataSubmission.html) for details, contact info, and instructions.

@@

| Gene name | Enrichment value (logFC) | p value |  AveExpr   |  FDR   |
|:----------|:------------------------:|:-------:|:----------:|:------:|
| GeneA     |           1.2            |  0.001  | 1.0x10^18^ | 0.0001 |
| GeneB     |           0.1            |   0.6   | 1.0x10^5^  | 0.005  |
| ...       |           ...            |   ...   |    ...     |  ...   |
: Example data table for smooth incorporation into Lipid Interactome. {#tbl-1 .striped}

@@

**Column descriptions:**

  -   *Gene name or Accession Number*
  -   *Enrichment value*: Fold change of Experimental versus Control fold-change, typically log2-transformed -- ideally the product of TMT or other quantitative mass spectrometry methods.
  -   *p value*: the results of an ANOVA or LIMMA analysis, corrected for multiple hypothesis testing
  -   *AveExpr*: An averaged value of the m/z ion intensity of the Experimental condition and Control condition used for the enrichment value (used to prepare MA plots for quality control)
  -   *FDR*: False-discovery rate of protein identification

@@

## Acknowledgements

-   Andre Nadler
-   Frank Stein
-   Per Haberkant

@@

{{< pagebreak >}}

## References

>>>>

copyright:
  holder: "Gaelen Guzman"
  year: 2025
authors:
  - name: "Gaelen Guzman"
    affiliations:
      - name:  "Oregon Health & Science University"
        address: 3222 SW Research Dr
        city: Portland
        state: OR
        postal-code: 97239
    orcid: 0000-0002-7696-6034
    corresponding: true
    email: gaelen.guzman@lipidinteractome.org
  - name: "Carsten Schultz"
    affiliations:
      - name:  "Oregon Health & Science University"
        address: 3222 SW Research Dr
        city: Portland
        state: OR
        postal-code: 97239
    orcid: 0000-0002-5824-2171
  - name: "Fikadu Tafesse"
    affiliations:
      - name:  "Oregon Health & Science University"
        address: 3222 SW Research Dr
        city: Portland
        state: OR
        postal-code: 97239
    orcid: 0000-0002-8575-4164

page-navigation: false
---

@@

### Data Submission & Contact

If you would like to submit a relevant dataset, see our [Data Submission Page](https://www.lipidinteractome.org/DataSubmission.html) for details, contact info, and instructions. For site feedback and additional questions, please email the site administrators at <Contact.Us@lipidinteractome.org>

@@

| Gene name | Enrichment value (logFC) | p value |  AveExpr   |  FDR   |
|:----------|:------------------------:|:-------:|:----------:|:------:|
| GeneA     |           1.2            |  0.001  | 1.0x10^18^ | 0.0001 |
| GeneB     |           0.1            |   0.6   | 1.0x10^5^  | 0.005  |
| ...       |           ...            |   ...   |    ...     |  ...   |
: **Example data table for smooth incorporation into Lipid Interactome.** ***Column descriptions:*** *Gene name*: The identifier for genes identified in the dataset (or Accession Number). *Enrichment value*: Fold change of Experimental versus Control fold-change, typically log2-transformed -- ideally the product of TMT or other quantitative mass spectrometry methods. *p value*: the results of an ANOVA or LIMMA analysis, corrected for multiple hypothesis testing. *AveExpr*: An averaged value of the m/z ion intensity of the Experimental condition and Control condition used for the enrichment value (used to prepare MA plots for quality control). *FDR*: False-discovery rate of protein identification. {#tbl-1 .striped}

@@

@@

## Contributions

GG designed, implemented, and wrote the majority of the LipidInteractome site and manuscript. FS wrote the data analysis section of the site, provided significant intellectual contribution to the site design, and edited the manuscript. AN provided substantial editing of the site and manuscript. CS and FT provided significant intellectual contribution and text editing of the site and manuscript, and oversaw the site development.

@@
